---
title: "Assignment 2: Coding Basics"
author: "Yuxiang Ren"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. 
seq(1, 100, 4) #generate sequence
sequence100 <- seq(1, 100, 4)  #assign a name

#2. 
resultmean <- mean(sequence100); resultmean  #compute mean, assign the result a name and show the result.
resultmedian <- median(sequence100); resultmedian  #compute median, assign the result a name and show the result.

#3. 
greater <- resultmean > resultmedian; greater #comparison, assign the result a name, and show the result.

```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
#5
Name <- c("Alex" ,"Reino", "Ellen", "Marcia") #vector type: character
Score <- c(78,90,32,66) #vector type: numeric
Pass <- c(TRUE, TRUE ,FALSE, TRUE)#vector type: logical

#7
df_name_score_pass <- data.frame("Name"= Name, "Score"= Score, "Pass" = Pass)



```

9.  QUESTION: How is this data frame different from a matrix?

> Answer:
Matrix only contains a single class of data. Similarly, each column in the data frame contains a single class. However, the different columns could be of different classes of data. 


10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement.

11. Apply your function to the vector with test scores that you created in number 5.

```{r}
#10

passing <- function(x){
  ifelse(x < 50, "FALSE", "TRUE")} #not added print in ifelse function, to avoid function print additional "FALSE" and "TRUE" before the result line. 

#11
passing(Score)

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 
'ifelse' worked. 
When using ‘if' and 'else' to operate the vector 'Score' with the condition (x<50), there is an error message: "the condition has length > 1". The reason is that in the 'if' and 'else' functions, the input x should be the only integer or object. Multiple values in the vector exceed the requirements of the condition. While in 'ifelse' function, it checks the condition for every element of a vector. Therefore, 'ifelse' worked.


