---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Yuxiang Ren"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
    
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on
generalized linear models.

```{r setup, include=FALSE}

#install.packages('formatR')

knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)

```

## Directions

1.  Rename this file `<FirstLast>_A06_GLMs.Rmd` (replacing `<FirstLast>`
    with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill
    each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code
    into a single PDF file.

## Set up your session

1.  Set up your session. Check your working directory. Load the
    tidyverse, agricolae and other needed packages. Import the *raw*
    NTL-LTER raw data file for chemistry/physics
    (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date
    objects.

2.  Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
library(tidyverse);library(agricolae);library(here);
library(ggplot2);library(lubridate)
here() #verify home directory
#raw data
Lake <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv",
                 stringsAsFactors = T)

Lake$sampledate <- mdy(Lake$sampledate)

#2
mytheme <- theme_bw(base_size = 14) +
  theme(axis.text=element_text(size=11,color="black"), 
        axis.title=element_text(size=13,color="black"),
        legend.title=element_text(size=13, color="black"))

```

## Simple regression

Our first research question is: Does mean lake temperature recorded
during July change with depth across all lakes?

3.  State the null and alternative hypotheses for this question: \>
    Answer: H0: Mean lake temperature recorded during July not change
    with depth across all lakes. Ha: Mean lake temperature recorded
    during July change with depth across all lakes.

4.  Wrangle your NTL-LTER dataset with a pipe function so that the
    records meet the following criteria:

-   Only dates in July.
-   Only the columns: `lakename`, `year4`, `daynum`, `depth`,
    `temperature_C`
-   Only complete cases (i.e., remove NAs)

5.  Visualize the relationship among the two continuous variables with a
    scatter plot of temperature by depth. Add a smoothed line showing
    the linear model, and limit temperature values from 0 to 35 °C. Make
    this plot look pretty and easy to read.

```{r scatterplot}
#4
Lake_temperature_depth <- Lake %>% filter(month(sampledate) == 7) %>% 
  select(lakename, year4, daynum, depth, temperature_C) %>% 
  drop_na()
  

#5
ggplot(Lake_temperature_depth, aes(x = depth, y = temperature_C))+
  geom_point(alpha = .4)+
  geom_smooth(method = "lm", col= "blue", se= FALSE)+
  labs(x = "Depth (m)", y = "Temperature (°C)")+
  ylim(0,35)+
  mytheme
  

```

6.  Interpret the figure. What does it suggest with regards to the
    response of temperature to depth? Do the distribution of points
    suggest about anything about the linearity of this trend?

> Answer: The scatter plot with the smoothed line suggests that there is
> a negative relationship between temperature and depth in the NTL-LTER
> lakes during July. As depth increases, the temperature tends to
> decrease. The distribution of points suggests that the relationship
> may not be perfectly linear. There appears to be a curvature to the
> relationship, with temperature decreasing more rapidly at shallower
> depths and the rate of decrease gradually levelling off as depth
> increases.

7.  Perform a linear regression to test the relationship and display the
    results

```{r linear.regression}
#7
lm_t_d <- lm(temperature_C ~ depth, data = Lake_temperature_depth)
summary(lm_t_d)

```

8.  Interpret your model results in words. Include how much of the
    variability in temperature is explained by changes in depth, the
    degrees of freedom on which this finding is based, and the
    statistical significance of the result. Also mention how much
    temperature is predicted to change for every 1m change in depth.

> Answer: The R-squared value of the linear regression model is 0.7387,
> which means that approximately 73.87% of the variability in
> temperature during July in the NTL-LTER lakes is explained by changes
> in depth. The degree of freedom for the linear regression model is
> 9726, which reflects the number of data points (9728) minus the number
> of estimated coefficients (2). This estimate of the relationship
> between depth and temperature is statistically significant, with a
> t-value of -165.8 and a p-value of \<2e-16, indicating that the
> probability of observing such a strong relationship by chance is very
> low. The result estimates that the mean temperature in the NTL-LTER
> lakes during July decreases by 1.94621 C for every 1-meter increase in
> depth.

------------------------------------------------------------------------

## Multiple regression

Let's tackle a similar question from a different approach. Here, we want
to explore what might the best set of predictors for lake temperature in
July across the monitoring period at the North Temperate Lakes LTER.

9.  Run an AIC to determine what set of explanatory variables (year4,
    daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables.

```{r temperature.model}
#9
year_day_dep_model <- lm(data = Lake_temperature_depth, temperature_C ~ year4 + daynum + depth)
step(year_day_dep_model)


#10
recommended_model <- lm(data = Lake_temperature_depth, temperature_C ~ year4 + daynum + depth)
summary(recommended_model)

```

11. What is the final set of explanatory variables that the AIC method
    suggests we use to predict temperature in our multiple regression?
    How much of the observed variance does this model explain? Is this
    an improvement over the model using only depth as the explanatory
    variable?

> Answer: The final set of explanatory variables includes year4, daynum
> and depth. They could explain approximately 74% of the observed
> variance. This multiple regression model is an improvement over the
> simple linear regression model due to having a lower AIC value and
> could explain more observed variance.

------------------------------------------------------------------------

## Analysis of Variance

12. Now we want to see whether the different lakes have, on average,
    different temperatures in the month of July. Run an ANOVA test to
    complete this analysis. (No need to test assumptions of normality or
    similar variances.) Create two sets of models: one expressed as an
    ANOVA models and another expressed as a linear model (as done in our
    lessons).

```{r anova.model}
#12
#ANOVA model
Anova_Lake_T_name <- aov(temperature_C ~ lakename, data = Lake_temperature_depth)
summary(Anova_Lake_T_name)
#Linear model
Lm_Lake_T_name <- lm(temperature_C ~ lakename, data = Lake_temperature_depth)
summary(Lm_Lake_T_name)
```

13. Is there a significant difference in mean temperature among the
    lakes? Report your findings.

> Answer: Yes, there is a significant difference in mean temperature
> among the lakes. First, the ANOVA model shows that the mean
> temperature in July varies significantly among the lakes. Second, the
> linear model shows that different lakes have different and significant
> intercepts and coefficients, which also indicates there is a
> significant difference in mean temperature among the lakes.

14. Create a graph that depicts temperature by depth, with a separate
    color for each lake. Add a geom_smooth (method = "lm", se = FALSE)
    for each lake. Make your points 50 % transparent. Adjust your y axis
    limits to go from 0 to 35 degrees. Clean up your graph to make it
    pretty.

```{r scatterplot.2}
#14.
Plot_14 <- ggplot(Lake_temperature_depth, aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  ylim(0, 35) +
  labs(x = "Depth (m)", y = "Temperature (C)", color = "Lake") +
  mytheme
print(Plot_14)

```

15. Use the Tukey's HSD test to determine which lakes have different
    means.

```{r tukey.test}
#15
Tukey_Lake_T_name <- TukeyHSD(Anova_Lake_T_name);Tukey_Lake_T_name

```

16.From the findings above, which lakes have the same mean temperature,
statistically speaking, as Peter Lake? Does any lake have a mean
temperature that is statistically distinct from all the other lakes?

> Answer: The mean temperature of Paul Lake might be the same as Lake
> Peter, with a confidence interval including zero. However, it's
> p-value is too high, making the result insignificant. There is no lake
> have a mean temperature that is statistically distinct from all the
> other lakes. Central Long Lake is different from most lakes but might
> be same as Crampton Lake due to it's confidence interval includiing
> zero.

17. If we were just looking at Peter Lake and Paul Lake. What's another
    test we might explore to see whether they have distinct mean
    temperatures?

> Answer: Two-sample t-test

18. Wrangle the July data to include only records for Crampton Lake and
    Ward Lake. Run the two-sample T-test on these data to determine
    whether their July temperature are same or different. What does the
    test say? Are the mean temperatures for the lakes equal? Does that
    match you answer for part 16?

```{r t-test}
Crampton_Ward <- filter(Lake_temperature_depth, lakename %in% c("Crampton Lake", "Ward Lake"))
t_test_Crampton_Ward <- t.test(temperature_C ~ lakename, data = Crampton_Ward)
t_test_Crampton_Ward
```

> Answer: The T-test shows that the estimated mean temperatures for the
> two lakes are different. However, due to the p-value > 0.05, it can
> not reject the null hypothesis that the mean temperatures for the two
> lakes are equal. Therefore, there is no statistically significant
> difference in mean temperature between Crampton Lake and Ward Lake in
> July.
